Probability of causal contact between interstellar civilizations through MonteCarlo simulations

### 2019-08-20 manuscript.tex
complete las llamadas a las figuras, falta ordenar

\begin{abstract}
%
The abundance of intelligent civilizations in the galaxy is a
longstanding question, which is often conceptualized as the problem
of the lack of received communication of the Fermi paradox.
%
Most efforts on the estimation of a number of intelligent
civilizations are centered on the Drake equation, although 
its factors are affected by large uncertainties, and it lacks a
temporal nature.
%
Here we present numerical simulations of stochastic processes of
emergence of civilizations with communication capabilities, and
the causal contacts among them.
%
We analyze the rate of causal contacts as a function of the mean
number of civilizations and the mean lifetime span distribution.
%
Our results indicate that, given the large distances involved, in
the odds to obtain a contact in a few years of monitoring, assuming a
perfect detection rate, are very rare.                                 
%
\end{abstract}

@@

\section{Motivations}
%\section{Motivations: Uncertainties and the non detection problem for ETI contact}
%{{{

@@

\section{Methods and working hypotheses}

%: accounting for causal contacts with 
%discrete event simulation process}\label{S_methods}
%{{{

@@

In what follows, we describe the experimental setup chosen to estimate
the probabilities of causal contacts and several derived quantities in
terms of three independent parameters, namely, the mean time span
between the appeareance of consecutive communicating civilizations
($\tau_a$), their mean lifetime ($\tau_s$), and the maximum distance a
signal can be detected by another civilization ($D_{max}$).
%
Intuitively, the shorter $\tau_a$ and the larger $\tau_s$ and
$D_{max}$, there is a greater probability of the existence of causal
contacts between pairs of intelligent civilizations in the Galaxy
(Table \ref{T_simu}).
%
The causal connection can be produced without the need of the two
civilizations being concurrent or active at the same time.
%
Each CETI will fill an annular region of the Galaxy, limited by two
surfaces.
%
The leading front, or surface of first contact (SFC) grows from the
central CETI until it reaches the distance $D_{max}$.
%
The trailing front, or surface of last contact (SLC) also grows from
the central CETI, with a delay with respect to the SFC equivalent to
the lifetime of the CETI, and produces an annular region that shrinks
until it reaches $D_{max}$, when all signal from the central CETI is
lost.
%
In the 2D simulation only the intersection of these spheres with the
plane of the galaxy is relevant, and produce the corresponding
circumferences.

@@

\hline
\end{tabular}
\caption{Definition of independent variables and fixed paramenters 
   that are part of the simulation.}
\label{T_simu}
\end{table}

@@

\setlength{\tabcolsep}{10pt}
\begin{table}
\centering
\begin{tabular}{cccc}
\hline
   fixed parameter & \multicolumn{3}{c}{adopted value}\hfill \\
\hline
   inner GHZ radius\hfill & \multicolumn{3}{c}{20000 ly}\\
   outer GHZ radius & \multicolumn{3}{c}{60000 ly}\\ \\
\hline
   variable parameter & min & max & N \\
\hline
   D$_{max}$ & 1000 ly & 20000 ly & 20\\
   $\tau_{a}$ & 1000 y & 20000 y & 20\\
   $\tau_{s}$ & 1000 y & 20000 y & 20\\
\hline
\end{tabular}
\caption{Adopted values for simulation parameters. Fixed parameters
   define the GHZ.  Variable parameters define the temporal structure
   of the process and the reach of messages.}
\label{T_parameters}
\end{table}

@@

\begin{figure}
   \centering
   \includegraphics[width=0.5\textwidth]{growingsphere.pdf}
   \caption{Schematic representation of the growing communicating
   sphere, over the space--time diagrams, where time is represented on
   the horizontal direction, and space is represented in the vertical
   direction.  In (a) the sphere is
   growing as the surface of first contact has not reached the maximum
   distance.  In (b) it has reached the maximum distance, so that it
   remains at the same size.  After a Doomsday event, the signals can
   still be observed, but the surface of last contact grows }
   \label{F_sphere}
\end{figure}

@@

We chose eight models that are on fairly oposed regions of the
explored parameter space, and cover short/long lifetimes, short/long
waiting times for the next CETI to appear in the Galaxy and
short/large range reach of the signal, including all possible
combinations.
%
The details of each model are summarized in the
Table~\ref{T_SelectedModels}.

In Fig.~\ref{F_res_1} we show the Empirical cumulative distributions
of the number contacts for CETIs in four different samplas (M5, M6,
M7, M8) with long range signal reach (D$_{max}$=10000). 
%
The lifetime is more determinant than the awakening rate in increasing
the number of contacts.
%
As expected, of the four samples, the sample M7, which corresponds to
a dense awakening in the timeline and a long lifetime, is the model
that maximizes the number of contacts, reaching a maximum of order 100
contacts for a single CETI.
%
This is considerable larger than any model with a shorter lifetime,
which produce a number of contacts of at most the order of ten
contacts per CETI.
%
A couple of comments are worth to mention about this plot.
%
First, it has a logarithmic scale on the x-axis, and second it is the
cumulative, not differential, empirical distribution.

@@

\subsection{Duration of contacts}
%{{{

@@

\subsection{Waiting time for a first contact}
%{{{

@@

In the Fig.~\ref{F_waiting} we show the fraction of the CETIs that
have made at least one contact as a funtion of the elapsed time since
the awakening, given that they make at least one contact on their
lifetime, for different models.
%
From a frequentist approach, the normalized cumulative counts are
related to the estimation of the probability of observing a given time
with no success.
%             
Both panels show the histograms of the waiting times for the first
contact, for all the CETIs that make at least one contact along their
lifetime.
%
We show separately the histograms for several values of $\tau_a$ and a
fixed value of $(\tau_s, D_{max})$, and for several values of $\tau_s$ and a
fixed value of $(\tau_a, D_{max})$.
%
For different values of $\tau_a$, the shape of the distribution is
similar, and the main difference given by the different number of
contacts on the different models.
%
Conversely, 
 For different values of $\tau_a$, the shape of the distribution is
similar, and the main difference given by the different number of
contacts on the different models.

@@

\subsection{Location within the Galaxy}
%{{{

On the Figure~\ref{F_radial} we show the fraction of CETIs that make
zero, one, two or three contacts, as a function of the galactocentric
distance.
%
For this model, which corresponds to XXX, the majority of CETIs do not
make any contact during all the time period in which they are active.
%
As expected, the fractions of CETIs making at least one contact
diminishes as the number of contacts rises.
%
The shape of this distribution is a result of the geometry of the GHZ.
%
In general, a closer position to the center of the Galaxy slightly favours 
the posibility of a contact.
%
This would indicate that the position of the Earth is suitable for the
odds of making a contact.

@@

\setlength{\tabcolsep}{10pt}
\begin{table*}
\centering
\begin{tabular}{ccccc}
\hline
Model subset & D$_{max}$ & $\tau_a$ & $\tau_s$ & description  \\
\hline
M1 & 40000 & [10000, 30000]   & [10000, 50000]   &dense, short lifetime\\
M2 & 40000 & [170000, 190000] & [10000, 50000]   &sparse, short lifetime\\
M3 & 40000 & [10000, 30000]   & [400000, 440000] &dense, long lifetime \\
M4 & 40000 & [170000, 190000] & [400000, 440000] &sparse, long lifetime\\

M5 & 80000 & [10000, 30000]   & [10000, 50000]   &dense, short lifetime\\
M6 & 80000 & [170000, 190000] & [10000, 50000]   &sparse, short lifetime\\
M7 & 80000 & [10000, 30000]   & [400000, 440000] &dense, long lifetime \\
M8 & 80000 & [170000, 190000] & [400000, 440000] &sparse, long lifetime\\

\hline
\end{tabular}
\caption{Selected models to analize the behevior of simulation outputs
   and their dependence on simulation outputs and their dependence on
   simulation parameters.}
\label{T_SelectedModels}
\end{table*}

\begin{figure} \centering
   %
   \includegraphics[width=0.5\textwidth]{F1.pdf}
   %
   \caption{Empirical cumulative distributions of the number contacts
   for CETIs in four different samplas (M5, M6, M7, M8) with long
   range signal reach (D$_{max}$=10000).
   %
   }
   % 
\label{F_res_1} \end{figure}

\begin{figure} % Distribution
   \centering
   \includegraphics[width=0.5\textwidth]{F2.pdf}
   \caption{Rate of CETIs that make no contact as a function of
   $\tau_a$ and $\tau_s$}
   \label{F_res_2}
\end{figure}

\begin{figure*} % 2D color plot
   \centering
   \includegraphics[width=0.3\textwidth]{m1d2.png}%
   \includegraphics[width=0.3\textwidth]{m1d3.png}%
   \includegraphics[width=0.3\textwidth]{m1d4.png}
   \caption{Rate of CETIs that never make contact (listening) as a
   function of $\tau_a$ and $\tau_s$, for 
   $D_{max}$=10000 (left panel),
   $D_{max}$=40000 (middle panel), and
   $D_{max}$=80000 (right panel)}
   \label{F_res_3}
\end{figure*}

\begin{figure*} % 2D color plot
   \centering
   \includegraphics[width=0.3\textwidth]{m2d2.png}%
   \includegraphics[width=0.3\textwidth]{m2d3.png}%
   \includegraphics[width=0.3\textwidth]{m2d4.png}
   \caption{Rate of CETIs that listen at the moment of the
   awakening, as a funcion of $\tau_a$ and $\tau_s$}
   \label{F_res_4}
\end{figure*}

@@

\begin{figure}
   \centering
   %\includegraphics[width=0.5\textwidth]{Figure_taua.png}%
   %\includegraphics[width=0.5\textwidth]{Figure_taus.png}
   \includegraphics[width=0.5\textwidth]{waiting_s+a_dif_ylog.pdf}
   %
   \caption{Histograms of the mean waiting times for the first contact, for
   several models.
   %
   Left panel shows the histograms for several values of $\tau_a$, and
   $\tau_s$ in the range 17000-24000 yr.
   %
   Right panel shows the histograms for several values of $\tau_s$, and
   $\tau_a$ in the range 68000-100000 yr.
   %
   The shape of the histograms does not change when $\tau_a$ is
   modified, but significantly changes when $\tau_s$ is modified.
   }
   \label{F_waiting}
\end{figure}

\begin{figure}
   \centering
   \includegraphics[width=0.5\textwidth]{radial.png}
   \caption{Fraction of CETIs making cero, one, two or three contacts as a function of
   galactocentric distance.}
   \label{F_res_5}
\end{figure}

>>>>

\begin{abstract}
%
The abundance of intelligent civilizations in the galaxy is a
longstanding question, which is often conceptualized as the problem
of the lack of received communication or the Fermi paradox.
%
Most efforts on the estimation of a number of intelligent
civilizations are centered on the Drake equation, although 
its factors are affected by large uncertainties, and it lacks a
temporal nature.
%
A different approach uses detailed numerical simulations of the galaxy
and recipes for the rates of the formation of stars and planets, and
even for the origin of life.
%
Here we present numerical simulations of stochastic processes of
emergence of civilizations with communication capabilities, based
on minimal asumptions, and analyze the causal connections among them.
%
The analysis of the rate of causal contacts as a function of the mean
number of civilizations, the mean lifetime span distribution and the
maximum distance a civilization can send signals is presented, and
used as a framework to discuss the spatial and temporal structure
of a populated galaxy within several different scenarios.
%
Our results indicate that, given the large distances involved, causal
contacts between civilizations are very rare.
%
Additionally, the odds to make a contact in a few years of monitoring, 
assuming a perfect detection rate, are low for most models, with the
exception of models which propose a galaxy densely populated with
long lived civilizations.
%
The probability of causal contacts increases with the increase of the 
mean lifetime of civilizations much more signifficantly than with
the increase in the mean number of active civilizations for a given
time window.
%
\end{abstract}

@@

\section{Motivations}\label{S_motivations}
%{{{

@@

\section{Methods and working hypotheses}\label{S_methods}
%{{{

@@

In what follows, we describe the experimental setup chosen to estimate
the probabilities of causal contacts and several derived quantities in
terms of three independent parameters, namely, the mean time span
between the appeareance of consecutive communicating civilizations
($\tau_a$), their mean lifetime ($\tau_s$), and the maximum distance a
signal can be detected by another civilization ($D_{max}$).
%
Intuitively, the shorter $\tau_a$ and the larger $\tau_s$ and
$D_{max}$, there is a greater probability of the existence of causal
contacts between pairs of intelligent civilizations in the Galaxy
(Table \ref{T_simu_hypotheses}).
%
The causal connection can be produced without the need of the two
civilizations being concurrent or active at the same time.
%
Each CETI will fill an annular region of the Galaxy, limited by two
surfaces.
%
The leading front, or surface of first contact (SFC) grows from the
central CETI until it reaches the distance $D_{max}$.
%
The trailing front, or surface of last contact (SLC) also grows from
the central CETI, with a delay with respect to the SFC equivalent to
the lifetime of the CETI, and produces an annular region that shrinks
until it reaches $D_{max}$, when all signal from the central CETI is
lost.
%
In the 2D simulation only the intersection of these spheres with the
plane of the galaxy is relevant, and produce the corresponding
circumferences.
%
We illustrate in the Fig.~\ref{F_growing_sphere} the schematic
representation of the growing communicating sphere, over the
space–time diagrams, where time is represented on the horizontal
direction, and space is represented in the vertical direction.
%
In the upper panel (a), the sphere is growing as the surface of first
contact has not reached the maximum distance.
%
In the middle panel (b) it has reached the maximum distance, so that
it remains at the same size.
%
After a Doomsday event (bottom panel, b), the signals can still be
observed, but the surface of last contact grows.
%
In the Fig.~\ref{F_abcd}, we show a schematic representation of two
emitters, E$_1$ and E$_2$, that reach each other at different times.
%
The time span for E$_i$ is (A$_i$ , D$_i$) , for $i = {1, 2}$.
%
Emitter $i$ can listen to emitter $j$ between C$_{ij}$ and B$_{ij}$.
%
The type and length of causal contact in both directions depend on the
distance and time lag between the awakening events, the maximum
distance that a signal can reach and the time period in which each
emitter is active.
%
In the Fig.~\ref{F_messages} we show a schematic representation of the
possible cases in which a CETI A can be in causal contact with a CETI
B which appears later.
%
The duration of the causal contact in one direction depends on several
factors, mainly $\Delta$t~A , $\Delta$t~C , and $\Delta$X.

@@

\hline
\end{tabular}
\caption{Definition of independent variables and fixed paramenters 
   that are part of the simulation.}
\label{T_simu_hypotheses}
\end{table}

@@

\setlength{\tabcolsep}{10pt}
\begin{table}
\centering
\begin{tabular}{cccc}
\hline
   fixed parameter & \multicolumn{3}{c}{adopted value}\hfill \\
\hline
   inner GHZ radius\hfill & \multicolumn{3}{c}{20000 ly}\\
   outer GHZ radius & \multicolumn{3}{c}{60000 ly}\\ \\
\hline
   variable parameter & min & max & N \\
\hline
   D$_{max}$ & 1000 ly & 20000 ly & 20\\
   $\tau_{a}$ & 1000 y & 20000 y & 20\\
   $\tau_{s}$ & 1000 y & 20000 y & 20\\
\hline
\end{tabular}
\caption{Adopted values for simulation parameters. Fixed parameters
   define the GHZ.  Variable parameters define the temporal structure
   of the process and the reach of messages.}
\label{T_range_parameters}
\end{table}

@@

\begin{figure}
   \centering
   \includegraphics[width=0.5\textwidth]{growingsphere.pdf}
   \caption{Schematic representation of the growing communicating
   sphere, over the space--time diagrams, where time is represented on
   the horizontal direction, and space is represented in the vertical
   direction.  In (a) the sphere is
   growing as the surface of first contact has not reached the maximum
   distance.  In (b) it has reached the maximum distance, so that it
   remains at the same size.  After a Doomsday event, the signals can
   still be observed, but the surface of last contact grows }
   \label{F_growing_sphere}
\end{figure}

@@

We chose eight models that are on fairly oposed regions of the
explored parameter space, and cover short/long lifetimes, short/long
waiting times for the next CETI to appear in the Galaxy and
short/large range reach of the signal, including all possible
combinations.
%
The details of each model are summarized in the
Table~\ref{T_selected_models}.

In Fig.~\ref{F_number_of_contacts} we show the Empirical cumulative
distributions of the number contacts for CETIs in four different
samples (M5, M6, M7, M8) with long range signal reach
(D$_{max}$=10000). 
%
The lifetime is more determinant than the awakening rate in increasing
the number of contacts.
%
As expected, of the four samples, the sample M7, which corresponds to
a dense awakening in the timeline and a long lifetime, is the model
that maximizes the number of contacts, reaching a maximum of order 100
contacts for a single CETI.
%
This is considerable larger than any model with a shorter lifetime,
which produce a number of contacts of at most the order of ten
contacts per CETI.
%
A couple of comments are worth to mention about this plot.
%
First, it has a logarithmic scale on the x-axis, and second it is the
cumulative, not differential, empirical distribution.

@@

\subsection{Multiplicity and duration of contacts}\label{SS_multiplicity}
%{{{

@@

\subsection{Waiting time for a first contact}\label{SS_waiting}
%{{{

@@

In the Fig.~\ref{F_waiting_time} we show the fraction of the CETIs
that have made at least one contact as a funtion of the elapsed time
since the awakening, given that they make at least one contact on
their lifetime, for different models.
%
From a frequentist approach, the normalized cumulative counts are
related to the estimation of the probability of observing a given time
with no success.
%             
Both panels show the histograms of the waiting times for the first
contact, for all the CETIs that make at least one contact along their
lifetime.
%
We show separately the histograms for several values of $\tau_a$ and a
fixed value of $(\tau_s, D_{max})$, and for several values of $\tau_s$
and a fixed value of $(\tau_a, D_{max})$.
%
For different values of $\tau_a$, the shape of the distribution is
similar, and the main difference given by the different number of
contacts on the different models.
%
Conversely, for different values of $\tau_a$, the shape of the
distribution is similar, and the main difference given by the
different number of contacts on the different models. 
%
We show the 2D maps of the rate of CETIs that never make contact and
the rate of CETIs that make contact at the moment of the awakening,
inf the Figs.~\ref{F_never_contact} and \ref{F_C_at_A}, respectively,
as a function of $\tau_s$ and $\tau_s$.
%
In the Fig.~\ref{F_waiting_for_1C} we show the histograms of the mean
waiting times for the first contact, for several models.
%
Left panel shows the histograms for several values of $\tau_a$, and
$\tau_s$ in the range 17000-24000 yr.
%
Right panel shows the histograms for several values of $\tau_s$, and
$\tau_a$ in the range 68000-100000 yr.
%
The shape of the histograms does not change when $\tau_a$ is modified,
but significantly changes when $\tau_s$ is modified.

@@

\subsection{Location within the Galaxy}\label{SS_location}
%{{{

On the Figure~\ref{F_radial_frac} we show the fraction of CETIs that make
zero, one, two or three contacts, as a function of the galactocentric
distance.
%
For this model, which corresponds to XXX, the majority of CETIs do not
make any contact during all the time period in which they are active.
%
As expected, the fractions of CETIs making at least one contact
diminishes as the number of contacts rises.
%
The shape of this distribution is a result of the geometry of the GHZ.
%
In general, a closer position to the center of the Galaxy slightly favours 
the posibility of a contact.
%
This would indicate that the position of the Earth is suitable for the
odds of making a contact.

@@

\setlength{\tabcolsep}{10pt}
\begin{table*}
\centering
\begin{tabular}{ccccc}
\hline
Model subset & D$_{max}$ & $\tau_a$ & $\tau_s$ & description  \\
\hline
M1 & 40000 & [10000, 30000]   & [10000, 50000]   &dense, short lifetime\\
M2 & 40000 & [170000, 190000] & [10000, 50000]   &sparse, short lifetime\\
M3 & 40000 & [10000, 30000]   & [400000, 440000] &dense, long lifetime \\
M4 & 40000 & [170000, 190000] & [400000, 440000] &sparse, long lifetime\\
%
M5 & 80000 & [10000, 30000]   & [10000, 50000]   &dense, short lifetime\\
M6 & 80000 & [170000, 190000] & [10000, 50000]   &sparse, short lifetime\\
M7 & 80000 & [10000, 30000]   & [400000, 440000] &dense, long lifetime \\
M8 & 80000 & [170000, 190000] & [400000, 440000] &sparse, long lifetime\\
%
\hline
\end{tabular}
\caption{Selected models to analize the behevior of simulation outputs
   and their dependence on simulation outputs and their dependence on
   simulation parameters.}
\label{T_selected_models}
\end{table*}

\begin{figure} \centering
   %
   \includegraphics[width=0.5\textwidth]{F1.pdf}
   %
   \caption{Empirical cumulative distributions of the number contacts
   for CETIs in four different samples (M5, M6, M7, M8) with long
   range signal reach (D$_{max}$=10000).
   %
   }
   % 
\label{F_number_of_contacts} \end{figure}

\begin{figure} % Distribution
   \centering
   \includegraphics[width=0.5\textwidth]{F2.pdf}
   \caption{Rate of CETIs that make no contact as a function of
   $\tau_a$ and $\tau_s$}
   \label{F_waiting_time}
\end{figure}

\begin{figure*} % 2D color plot
   \centering
   \includegraphics[width=0.3\textwidth]{m1d2.png}%
   \includegraphics[width=0.3\textwidth]{m1d3.png}%
   \includegraphics[width=0.3\textwidth]{m1d4.png}
   \caption{Rate of CETIs that never make contact (listening) as a
   function of $\tau_a$ and $\tau_s$, for 
   $D_{max}$=10000 (left panel),
   $D_{max}$=40000 (middle panel), and
   $D_{max}$=80000 (right panel)}
   \label{F_never_contact}
\end{figure*}

\begin{figure*} % 2D color plot
   \centering
   \includegraphics[width=0.3\textwidth]{m2d2.png}%
   \includegraphics[width=0.3\textwidth]{m2d3.png}%
   \includegraphics[width=0.3\textwidth]{m2d4.png}
   \caption{Rate of CETIs that listen at the moment of the
   awakening, as a funcion of $\tau_a$ and $\tau_s$}
   \label{F_C_at_A}
\end{figure*}

@@

\begin{figure}
   %
   \centering
   %
   \includegraphics[width=0.5\textwidth]{waiting_s+a_dif_ylog.pdf}
   %
   \caption{Histograms of the mean waiting times for the first contact, for
   several models.
   %
   Left panel shows the histograms for several values of $\tau_a$, and
   $\tau_s$ in the range 17000-24000 yr.
   %
   Right panel shows the histograms for several values of $\tau_s$, and
   $\tau_a$ in the range 68000-100000 yr.
   %
   The shape of the histograms does not change when $\tau_a$ is
   modified, but significantly changes when $\tau_s$ is modified.
   }
   %
   \label{F_waiting_for_1C}
   %
\end{figure}

\begin{figure}
   %
   \centering
   %
   \includegraphics[width=0.5\textwidth]{radial.png}
   %
   \caption{Fraction of CETIs making cero, one, two or three contacts
   as a function of galactocentric distance.}
   %
   \label{F_radial_frac}
   %
\end{figure}
